cubble An R Package for Structuring Spatio-temporal Data

### 2022-04-21 1-intro.Rmd
introduction done

Spatio-temporal data has a spatial component referring to the location of each observation, a temporal component that is recorded at regular or irregular time intervals, and multiple variables measured at each spatial and temporal value. The analysis of spatio-temporal data can fix the time to explore the spatial distribution, fix the spatial location to explore temporal trends, or dynamically explore the space and time simultaneously. For these tasks to be carried out easily, there is a need for a spatio-temporal data representation that underpins various operations.

The *Representing data* section of the CRAN task view [ref] lists existing spatio-temporal data representation. Among them, @spacetime proposes four layouts: full grid layout, sparse grid layout, irregular layout, and trajectory. These layouts are implemented in the \pkg{spacetime} package with underlying class \pkg{sp} [@sp] and \pkg{xts} [@xts] for handling spatial and temporal variables. The \pkg{stars} package has been a recent implementation to handle spatio-temporal data in an array. The package depends on the \pkg{sf} package and has allowed more access to tidyverse verbs in the workflow.

What remains to be difficult when wrangling spatio-temporal data is that the raw data is not naturally structured in any one of the representation above and the first task is to cast the data into a class. Also, in data collection, spatio-temporal data is less often collected as a whole piece but fragmented spatial and temporal data for analysts to ensemble. For example, map data records the shape of a collection of area of interest is provided by official government website; geo-scientific data records the longitude and latitude coordinate of locations in the area along with other meta data relates to the location; and temporal variables indexed by both location and time. For spatio-temporal data analysis, merging all these sources can be challenging and joining these tables can go wrong because of the slightest unmatched of the linking variable from different sources.

Another issue is how the concept of tidy data [@tidydata] should be applied to spatio-temporal data. The tidy data concept advocates the data to be shaped into 1) one row per observation; 2) one column per variable; and 3) one type of data per table. The long form data is preferred over wide data given the downstream software i.e. \pkg{dplyr} and \pkg{ggplot2} for data wrangling and visualisation. However, the long form can be inefficient to store feature geometries, especially for large multipolygon over long time period. This poses the question on how to arrange spatial and temporal variables that makes analysing spatio-temporal data easier.

This paper presents a new \proglang{R} package, \pkg{cubble}, to [overcome the problem] with organising variables in different levels when working with spatio-temporal data. The data structure focuses on connecting to the upstream data collection and is suitable for temporal data collected at the fixed location over time. Among the four spacetime layouts in @spacetime, it can be full grid layout, sparse grid layout, or irregular layout, but trajectory is not within the scope of this work. In the package a new data structure, also called \code{cubble}, is proposed to organise spatial and temporal variables as two forms of a single data object so that they can be wrangled separately while synchronised as a whole piece. With this new data object, analysts can spend less time writing codes to organise spatial and temporal variables and focus more on the exploratory data analysis itself. The software is available from the Comprehensive R Archive Network (CRAN) at [CRAN link].

The rest of the paper is divided as follows: Section \@ref(cube) discusses the spatio-temporal cube to conceptualise spatio-temporal data. Section \@ref(cubble) presents the main design and functionality of \pkg{cubble}. Section \@ref(others) explains how cubble deals with more advanced considerations including data with hierarchical structure, data matching, how cubble fits with existing static and interactive visualisation tools, and spatio-temporal data transformation. Section \@ref(examples) uses Australia weather station data and river level data as examples to demonstrate the use of \pkg{cubble}. An example on how \pkg{cubble} handles NetCDF data is also provided. Section \@ref(conclude) concludes the paper.

>>>>

Spatio-temporal data has a spatial component referring to the location of each observation, a temporal component that is recorded at regular or irregular time intervals, and may have multiple variables measured at each spatial and temporal value. With spatio-temporal data, one can fix the time to explore the spatial distribution, fix the spatial location to explore temporal trends, or dynamically explore the space and time simultaneously. Underlying all these tasks is a data representation that enables the spatio-temporal data to be sliced and diced in different ways.

The SpatioTemproal CRAN task view [@ctvspatiotemporal] has a section *Representing data* that lists existing spatio-temporal data representation in \proglang{R}. Among them, @spacetime proposes four spatio-temporal layouts: full grid layout, sparse grid layout, irregular layout, and trajectory. These layouts are implemented in the \pkg{spacetime} [@spacetime] package with the underlying spatial and temporal class: \pkg{sp} [@sp] and \pkg{xts} [@xts]. The \pkg{stars} [@stars] package has been a recent implementation to handle both raster and vector data using spatio-temporal arrays. It interfaces with \pkg{sf} [@sf], which adopts the simple features representation of spatial data, and \pkg{tidyverse} [@tidyverse], a suite of tools to wrangle and visualise data in \proglang{R}.

Still, the data representation for spatio-temporal data can be further improved and there are two reasons for this. Firstly, the raw data sourced in the wild is less often presented in any one of the representations above, and fitting the raw data into one of the representations can sometimes be not easy. More often, spatio-temporal data is collected in separate 2D tables and analysts need to ensemble them into a whole piece before exploring the data. Examples of components of spatio-temporal data can be 1) map data records the shape of a collection of area of interest is provided by the official government websites; 2) geo-scientific data records the longitude and latitude coordinates of locations in the area along with other meta data relates to the location, and; 3) temporal variables indexed by both location and time.

<!-- For spatio-temporal data analysis, merging all these sources can be challenging and joining these tables can go wrong because of the slightest unmatched of the linking variable from different sources.  -->

The other reason relates to how the concept of tidy data [@tidydata] should be applied to spatio-temporal data. The tidy data concept advocates the data to be shaped into 1) one row per observation; 2) one column per variable, and; 3) one type of data per table. The long form data is preferred over wide data given the downstream software i.e. \pkg{dplyr} [@dplyr] and \pkg{ggplot2} [@ggplot2] for data wrangling and visualisation. However, the long form can be inefficient to store feature geometries, especially for large multipolygon over daily or sub-daily data over years, which are extensively collected and handled in time series analysis. This poses the question of how to arrange spatial and temporal variables that makes analysing spatio-temporal data easier.

This paper presents a new \proglang{R} package, \pkg{cubble}, for working with a new spatio-temporal data representation, which addresses the two issues mentioned above. In the package, a new data structure, also called \code{cubble}, is proposed to organise spatial and temporal variables as two forms of a single data object so that they can be wrangled separately while synchronised as a whole piece. Among the four spacetime layouts in @spacetime, \pkg{cubble} can be applied to full grid layout, sparse grid layout, or irregular layout, but not trajectory, which is outside the scope of this work. The software is available from the Comprehensive R Archive Network (CRAN) at [CRAN link].

The rest of the paper is divided as follows: Section \@ref(cube) discusses the spatio-temporal cube to conceptualise spatio-temporal data. Section \@ref(cubble) presents the main design and functionality of \pkg{cubble}. Section \@ref(others) explains how cubble deals with more advanced considerations including data with the hierarchical structure, data matching, how cubble fits with existing static and interactive visualisation tools, and spatio-temporal data transformation. Section \@ref(examples) uses Australia weather station data and river level data as examples to demonstrate the use of \pkg{cubble}. An example of how \pkg{cubble} handles NetCDF data is also provided. Section \@ref(conclude) concludes the paper.

### 2022-04-21 arxiv/cubble-arxiv.tex
introduction done

Spatio-temporal data has a spatial component referring to the location of each observation, a temporal component that is recorded at regular or irregular time intervals, and may have multiple variables measured at each spatial and temporal value. The analysis of spatio-temporal data can fix the time to explore the spatial distribution, fix the spatial location to explore temporal trends, or dynamically explore the space and time simultaneously. For these tasks to be carried out easily, there is a need for a spatio-temporal data representation that underpins various operations.

The \emph{Representing data} section of the SpatioTemproal CRAN task view (\protect\hyperlink{ref-ctvspatiotemporal}{Edzer Pebesma 2022}) lists existing spatio-temporal data representation. Among them, E. Pebesma (\protect\hyperlink{ref-spacetime}{2012}) proposes four layouts: full grid layout, sparse grid layout, irregular layout, and trajectory. These layouts are implemented in the \pkg{spacetime} (\protect\hyperlink{ref-spacetime}{E. Pebesma 2012}) package with the underlying spatial and temporal class: \pkg{sp} (\protect\hyperlink{ref-sp}{E. Pebesma and Bivand 2005}) and \pkg{xts} (\protect\hyperlink{ref-xts}{Ryan and Ulrich 2020}). The \pkg{stars} (\protect\hyperlink{ref-stars}{E. Pebesma 2021}) package has been a recent implementation to handle both raster and vector spatio-temporal data using array. It depends on the newer spatial package \pkg{sf} (\protect\hyperlink{ref-sf}{E. J. Pebesma 2018}) and interfaces with the widely-used tidyverse verbs.

{[}till here{]} One issue with wrangling spatio-temporal data is that the raw data does not naturally structured in any one of the representation above. Also, in data collection, spatio-temporal data is less often collected as a whole piece but fragmented spatial and temporal data for analysts to ensemble. For example, map data records the shape of a collection of area of interest is provided by official government website; geo-scientific data records the longitude and latitude coordinate of locations in the area along with other meta data relates to the location; and temporal variables indexed by both location and time. For spatio-temporal data analysis, merging all these sources can be challenging and joining these tables can go wrong because of the slightest unmatched of the linking variable from different sources.

Another issue is how the concept of tidy data (\protect\hyperlink{ref-tidydata}{Wickham 2014}) should be applied to spatio-temporal data. The tidy data concept advocates the data to be shaped into 1) one row per observation; 2) one column per variable; and 3) one type of data per table. The long form data is preferred over wide data given the downstream software i.e.~\pkg{dplyr} (\protect\hyperlink{ref-dplyr}{Wickham et al. 2022}) and \pkg{ggplot2} (\protect\hyperlink{ref-ggplot2}{Wickham 2016}) for data wrangling and visualisation. However, the long form can be inefficient to store feature geometries, especially for large multipolygon over long time period. This poses the question on how to arrange spatial and temporal variables that makes analysing spatio-temporal data easier.

This paper presents a new \proglang{R} package, \pkg{cubble}, to {[}overcome the problem{]} with organising variables in different levels when working with spatio-temporal data. The data structure focuses on connecting to the upstream data collection and is suitable for temporal data collected at the fixed location over time. Among the four spacetime layouts in E. Pebesma (\protect\hyperlink{ref-spacetime}{2012}), it can be full grid layout, sparse grid layout, or irregular layout, but trajectory is not within the scope of this work. In the package a new data structure, also called \code{cubble}, is proposed to organise spatial and temporal variables as two forms of a single data object so that they can be wrangled separately while synchronised as a whole piece. With this new data object, analysts can spend less time writing codes to organise spatial and temporal variables and focus more on the exploratory data analysis itself. The software is available from the Comprehensive R Archive Network (CRAN) at {[}CRAN link{]}.

The rest of the paper is divided as follows: Section \ref{cube} discusses the spatio-temporal cube to conceptualise spatio-temporal data. Section \ref{cubble} presents the main design and functionality of \pkg{cubble}. Section \ref{others} explains how cubble deals with more advanced considerations including data with hierarchical structure, data matching, how cubble fits with existing static and interactive visualisation tools, and spatio-temporal data transformation. Section \ref{examples} uses Australia weather station data and river level data as examples to demonstrate the use of \pkg{cubble}. An example on how \pkg{cubble} handles NetCDF data is also provided. Section \ref{conclude} concludes the paper.

@@

>>>>

Spatio-temporal data has a spatial component referring to the location of each observation, a temporal component that is recorded at regular or irregular time intervals, and may have multiple variables measured at each spatial and temporal value. With spatio-temporal data, one can fix the time to explore the spatial distribution, fix the spatial location to explore temporal trends, or dynamically explore the space and time simultaneously. Underlying all these tasks is a data representation that enables the spatio-temporal data to be sliced and diced in different ways.

The SpatioTemproal CRAN task view (\protect\hyperlink{ref-ctvspatiotemporal}{Edzer Pebesma 2022}) has a section \emph{Representing data} that lists existing spatio-temporal data representation in \proglang{R}. Among them, E. Pebesma (\protect\hyperlink{ref-spacetime}{2012}) proposes four spatio-temporal layouts: full grid layout, sparse grid layout, irregular layout, and trajectory. These layouts are implemented in the \pkg{spacetime} (\protect\hyperlink{ref-spacetime}{E. Pebesma 2012}) package with the underlying spatial and temporal class: \pkg{sp} (\protect\hyperlink{ref-sp}{E. Pebesma and Bivand 2005}) and \pkg{xts} (\protect\hyperlink{ref-xts}{Ryan and Ulrich 2020}). The \pkg{stars} (\protect\hyperlink{ref-stars}{E. Pebesma 2021}) package has been a recent implementation to handle both raster and vector data using spatio-temporal arrays. It interfaces with \pkg{sf} (\protect\hyperlink{ref-sf}{E. J. Pebesma 2018}), which adopts the simple features representation of spatial data, and \pkg{tidyverse} (\protect\hyperlink{ref-tidyverse}{Wickham et al. 2019}), a suite of tools to wrangle and visualise data in \proglang{R}.

Still, the data representation for spatio-temporal data can be further improved and there are two reasons for this. Firstly, the raw data sourced in the wild is less often presented in any one of the representations above, and fitting the raw data into one of the representations can sometimes be not easy. More often, spatio-temporal data is collected in separate 2D tables and analysts need to ensemble them into a whole piece before exploring the data. Examples of components of spatio-temporal data can be 1) map data records the shape of a collection of area of interest is provided by the official government websites; 2) geo-scientific data records the longitude and latitude coordinates of locations in the area along with other meta data relates to the location, and; 3) temporal variables indexed by both location and time.

The other reason relates to how the concept of tidy data (\protect\hyperlink{ref-tidydata}{Wickham 2014}) should be applied to spatio-temporal data. The tidy data concept advocates the data to be shaped into 1) one row per observation; 2) one column per variable, and; 3) one type of data per table. The long form data is preferred over wide data given the downstream software i.e.~\pkg{dplyr} (\protect\hyperlink{ref-dplyr}{Wickham et al. 2022}) and \pkg{ggplot2} (\protect\hyperlink{ref-ggplot2}{Wickham 2016}) for data wrangling and visualisation. However, the long form can be inefficient to store feature geometries, especially for large multipolygon over daily or sub-daily data over years, which are extensively collected and handled in time series analysis. This poses the question of how to arrange spatial and temporal variables that makes analysing spatio-temporal data easier.

This paper presents a new \proglang{R} package, \pkg{cubble}, for working with a new spatio-temporal data representation, which addresses the two issues mentioned above. In the package, a new data structure, also called \code{cubble}, is proposed to organise spatial and temporal variables as two forms of a single data object so that they can be wrangled separately while synchronised as a whole piece. Among the four spacetime layouts in E. Pebesma (\protect\hyperlink{ref-spacetime}{2012}), \pkg{cubble} can be applied to full grid layout, sparse grid layout, or irregular layout, but not trajectory, which is outside the scope of this work. The software is available from the Comprehensive R Archive Network (CRAN) at {[}CRAN link{]}.

The rest of the paper is divided as follows: Section \ref{cube} discusses the spatio-temporal cube to conceptualise spatio-temporal data. Section \ref{cubble} presents the main design and functionality of \pkg{cubble}. Section \ref{others} explains how cubble deals with more advanced considerations including data with the hierarchical structure, data matching, how cubble fits with existing static and interactive visualisation tools, and spatio-temporal data transformation. Section \ref{examples} uses Australia weather station data and river level data as examples to demonstrate the use of \pkg{cubble}. An example of how \pkg{cubble} handles NetCDF data is also provided. Section \ref{conclude} concludes the paper.

@@

\leavevmode\vadjust pre{\hypertarget{ref-tidyverse}{}}%
Wickham, Hadley, Mara Averick, Jennifer Bryan, Winston Chang, Lucy D'Agostino McGowan, Romain François, Garrett Grolemund, et al. 2019. {``Welcome to the {tidyverse}.''} \emph{Journal of Open Source Software} 4 (43): 1686. \url{https://doi.org/10.21105/joss.01686}.
